On the Minimum Core Mass for Giant Planet Formation

### 2013-10-03 tex/draft/draft4.tex
revised paper I intro

Current theories of giant planet formation postulate that these planets form either through core accretion or gravitational instability (GI). In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope. GI theories postulate that an instability in the protoplanetary disk leads to the fragmentation of the disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.

@@

Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges. Even though GI is the preferred mechanism at large distances (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}. Moreover, core accretion times are thought to be too long in the outer disk, largely due to the time it takes to accrete the solid core. Some models, however, allow for rapid accretion of solids even at large separations (e.g., \citealt{dones93}, \citealt{kenyon09}, \citealt{lambrechts12}). We thus consider the timescale constraints imposed by the accretion of the gaseous atmosphere before the protoplanetary disk dissipates.

@@

Several types of core accretion models have been studied previously. Analytic models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that the luminosity of the gaseous envelope is entirely supplied by planetesimal accretion. These studies consider that the atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope. The core and the atmosphere grow simultaneously, and for a fixed planetesimal accretion rate every core mass maps uniquely to one atmosphere mass. As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences; this process is called the ``core accretion instability''. Thus, there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.  Typically, the critical core mass is found to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).

@@

However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and the envelope. In these studies, the atmosphere luminosity is supplied not only by planetesimal accretion, but also by the Kelvin-Helmholtz (KH) contraction of the envelope itself. \citet{pollack96} find that planetesimal accretion significantly slows down before a massive gaseous envelope can form, due to depletion of the planet's feeding zone, and gas accretion starts to dominate the atmosphere evolution. In this situation, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts . This phase of slow gas accretion is found to be much longer than the formation of the solid core (\citealt{pollack96}, \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction rather than planetesimal accretion.

Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to consider atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}). This is the regime we assume in this work. Specifically, we study the atmosphere evolution of giant planets under the assumption that the solid core is fully grown and thus the planetesimal accretion rate is negligible. As a result, the luminosity of the gaseous envelope is solely supplied by atmospheric contraction. The envelope is not in a steady state, but rather it accretes gas as it cools radiatively. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and a rapid phase of runaway gas accretion commences. We define the time required for the atmosphere to grow to this stage as the \textit{crossover time}. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the \textit{critical core mass}.

>>>>

Current theories of giant planet formation postulate that these planets form either through core accretion or gravitational instability (GI). In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope. GI theories postulate that giant planets can be formed due to the fragmentation of the protoplanetary disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.

@@

Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges. Even though GI works better at large distances than in the inner disk (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}. On the other hand, core accretion times are thought to be too long in the outer disk, largely due to the time it takes to accrete the solid core. Some models, however, allow for rapid accretion of solids even at large separations (e.g., \citealt{dones93}, \citealt{kenyon09}, \citealt{lambrechts12}). We thus consider the timescale constraints imposed by the accretion of the gaseous atmosphere before the protoplanetary disk dissipates.

@@

Several types of core accretion models have been studied previously. In static models, which are mostly analytic, the accretion rate of planetesimals regulates the growth of the atmosphere  (e.g., \citealt{stevenson82}, \citealt{rafikov06}). The luminosity of the gaseous envelope is thus entirely supplied by planetesimal accretion. These studies consider that the atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope. For a fixed planetesimal accretion rate, the atmosphere mass grows non-linearly as the core mass increases. As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and the static model breaks down. Unstable collapse occurs if the core mass further increases, a process known as the ``core accretion instability''. The maximum core mass for which the atmosphere is still in hydrostatic equilibrium is defined as the ``critical core mass''.  Typically, the critical core mass is found to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).

@@

Full time dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope. As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively. These evolutionary models for core and atmosphere growth allow for the planetesimal accretion rate to be directly calculated as a function of time rather than prescribed, as it is the case in the static models. \citet{pollack96} find that the planet growth can be separated into three relatively well defined phases. Phase 1 is dominated by the growth of the core due to runaway planetesimal accretion, and ends when the planet's feeding zone has been depleted of solids. At this stage, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts. This is phase 2, during which the envelope mass steadily increases, while core growth is no longer significant. Once the atmosphere mass becomes comparable to the core mass, growth is continually accelerated, leading to runaway gas accretion, i.e. phase 3. Phase 2 is found to be much longer than the other two phases (see also \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction.

%However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and the envelope. In these studies, the atmosphere luminosity is supplied not only by planetesimal accretion, but also by the Kelvin-Helmholtz (KH) contraction of the envelope itself. \citet{pollack96} find that planetesimal accretion significantly slows down before a massive gaseous envelope can form, due to depletion of the planet's feeding zone, and gas accretion starts to dominate the atmosphere evolution. In this situation, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts . This phase of slow gas accretion is found to be much longer than the formation of the solid core (\citealt{pollack96}, \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction rather than planetesimal accretion.

Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}). This regime is considered in this work. We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass needed to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas. We note that our models are a special case of the time dependent studies, with the planetesimal accretion rate set to zero.

% Specifically, we study the atmosphere evolution of giant planets under the assumption that the solid core is fully grown and thus the planetesimal accretion rate is negligible.

%As a result, the luminosity of the gaseous envelope is solely supplied by atmospheric contraction. The envelope is not in a steady state, but rather it accretes gas as it cools radiatively. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and a rapid phase of runaway gas accretion commences. We define the time required for the atmosphere to grow to this stage as the \textit{crossover time}. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the \textit{critical core mass}.
